---
title: "Basic analysis of ToothGrowth dataset"
author: "Yevgen Yampolskiy"
date: "01/23/2015"
output: html_document
---

`ToothGrowth` dataset describes the effect of vitamin C on tooth growth in 
guinea pigs.

The response is the length of odontoblasts (teeth) in each of `10` guinea pigs 
at each of three dose levels of Vitamin C (`0.5`, `1`, and `2 mg`) with each of two 
delivery methods (orange juice or ascorbic acid) (source: `?ToothGrowth`).

## Explore basic features
I found the most convenient to look at this data using the following `boxplot`:
```{r}
data(ToothGrowth)
boxplot(len ~ dose + supp, data = ToothGrowth)
```

From this plot we can assume that length increases as the dose increases, and 
that orange juice method is more effective for small concentrations. 
For concentration of `2 mg` method choice does not make much difference, 
although for asorbic acid we see higher variance. 

## Significance of results
For each concentration (dose), do we have significant difference in 
mean values?
```{r}
t.test(len ~ supp,  data=subset(ToothGrowth, dose == 0.5))
t.test(len ~ supp,  data=subset(ToothGrowth, dose == 1.0))
t.test(len ~ supp,  data=subset(ToothGrowth, dose == 2.0))
```
We see that for concentrations of `0.5 mg` and `1.0 mg` the `95%` confidence 
interval is above zero, and for concentration of `2.0 mg` confidence interval 
contains zero, which supports initial observation.

## Study assumptions
Fundamental assumption we are making is that guinea pigs for each group had 
equal qualities. For example, if orange juice and Vitamin C where given to 
guinea pigs of different breeds, or one group contained only male pigs while 
another group only female than non-homogenuity could be the driving factor. 
One way to archive homogenuity is to form groups *at random*.

To see that this is really the case it worth checking data source. 
In our case detailed description of the data set could be found in 
*Bliss (1952) The Statistics of Bioassay. Academic Press, pp. 499-500*:

**Example 9a**. A biological assay of the vitamin C activity of fresh orange juice 
may be used as a second example. The response of `10` guinea pigs on each of 
three doses of ascorbic acid (`S`) and of fresh orange juice (`U`) during a 
six-week test period was measured by Crampton (26) from the average length 
of the odontoblasts in each animal (p. 257), with the results given in 
`Table XXVIII`. Five of the animals at each treatment combination were males 
and five were females but except for this sex restriction the guinea pigs were 
assigned to doses entirely at random. The assay could be considered, 
therefore, as forming a `2 X 3 X 2` factorial experiment and has been so 
analyzed. The complete analysis (not given here) showed that the effect of sex 
could be neglected, so that in the interests of simplicity the data will be 
treated here as a `2 X 3` assay with `10` responses at each dose.

*Table skipped*

The fresh orange juice or unknown was analyzed chemically for its ascorbic 
acid content and on this basis fed without alteration at the same three 
levels of ascorbic acid as the standard. The ascorbic acid standard was 
administered at `0.5`, `1.0` and `2.0 mg` per day in a mixture of corn starch 
and corn meal. Hence the objective of the assay was to determine whether 
under these conditions the antiscorbutic activity of the orange juice was 
greater than would be predicted from its content of ascorbic acid as determined
chemically in the present experiment.


